---
title: "14-Day COVID-19 Forecast for Utah"
date: last-modified
execute:
  echo: false
---

```{r}
#| label: run-the-forecast
# This will render the methodology file, that generates the forecast
# Since we have `cache: true` in `methodology.qmd`, computations are
# cached and will not be re-run if the file has not changed.
system2("quarto", c("render", "methodology.qmd"))
```

```{r}
#| label: load-forecasts

source("functions.R")
fn <- sprintf(
  "forecast_data_%s.rds",
  format(Sys.time(), "%Y-%m-%d")
)

forecast <- readRDS(fn)
```

## Introduction

Using case count data published by the Utah DHHS and the [epiworldR](https://github.com/UofUEpiBio/epiworldR) package, we calibrate and run an SIR connected model to generate a COVID-19 forecast for Utah.
The 14-day forecast updates automatically each week.

## COVID-19 Forecast
Here is the forecast mean along with the 50% and 95% confidence intervals.
The actual case counts are plotted in black, while the forecast is plotted in blue.

```{r}
#| label: plot-forecast
with(forecast, plot_forecast(forecast_dist, covid_data))
```

## Acknowledgements
This forecast was developed jointly by Andrew Pulsipher and George Vega Yon as part of the EpiForeSITE project.
