---
title: "14-Day COVID-19 Forecast for Utah"
date: last-modified
execute:
  echo: false
---

```{r}
#| label: run-the-forecast
# This will render the methodology file, that generates the forecast
# Since we have `cache: true` in `methodology.qmd`, computations are
# cached and will not be re-run if the file has not changed.
quarto::quarto_render("methodology.qmd")
```

```{r}
#| label: load-forecasts

source("functions.R")

# Find all forecast data files matching the pattern
files <- list.files(pattern = "^forecast_data_\\d{4}-\\d{2}-\\d{2}\\.rds$")
if (length(files) == 0) {
  stop("No forecast data files found. Please render methodology.qmd first.")
}
# Extract dates from filenames and find the most recent
file_dates <- as.Date(sub("forecast_data_(\\d{4}-\\d{2}-\\d{2})\\.rds", "\\1", files))
most_recent_file <- files[which.max(file_dates)]
forecast <- readRDS(most_recent_file)
```

## Introduction

Using case count data published by the Utah DHHS and the [epiworldR](https://github.com/UofUEpiBio/epiworldR) package, we calibrate and run an SIR connected model to generate a COVID-19 forecast for Utah.
The 14-day forecast updates automatically each week.

## COVID-19 Forecast
Here is the forecast mean along with the 50% and 95% confidence intervals.
The actual case counts are plotted in black, while the forecast is plotted in blue.

```{r}
#| label: plot-forecast
with(forecast, plot_forecast(forecast_dist, covid_data))
```

## Acknowledgements
This forecast was developed jointly by Andrew Pulsipher and George Vega Yon as part of the EpiForeSITE project.
