---
title: "Methodology"
date: last-modified
execute:
  echo: false
---

```{r}
#| label: load-forecast-script
source("covid-forecast.R")
```

## Introduction

Using case data published by the Utah DHHS and the [epiworldR](https://github.com/UofUEpiBio/epiworldR) package, we calibrate and run an SIR connected model to generate a 14-day COVID-19 forecast for Utah, which is published weekly.

## COVID-19 Cases in Utah (last 90 days)
Every week, Utah DHHS publishes COVID-19 surveillance data on their [Coronavirus Dashboard](https://coronavirus.utah.gov/case-counts/) which includes reported case counts for each day starting March 18, 2020.
Our forecast is calibrated on case counts from the last 90 days.

```{r}
#| label: plot-observed-data
#| fig-width: 7
#| fig-height: 3
#| fig-align: center
plot_covid_data(covid_data)
```

## Calibrating the Forecasting Model
We use epiworldR's implementation of Likelihood-Free Markhov Chain Monte Carlo (LFMCMC) to calibrate the SIR connected model (using ModelSIRCONN from epiworldR) by estimating the following model parameters:

* Recovery rate
* Transmission rates for each season (spring, summer, fall, winter)
* Contact rates for weekdays and weekends

The LFMCMC simulation runs for 6,000 iterations.
In each iteration, it does the following:

* Proposes a new set of parameter values
* Runs the model with 10,000 agents and the proposed parameters
* Compares the output to the UDHHS data and records the result

We note that while the population of Utah over 3 million, we only need to run our model with 10,000 agents.
This is because we don't expect more than 10,000 COVID-19 cases and the model automatically scales the contact rate to account for the difference between the model population and Utah's population.

When the simulation is finished, we use a burn-in period of n = 2,000 (33% of simulation iterations).
The epiworldR results printout (below) shows the mean parameter/statistic value, the 95% credible interval (in `[ ]`), and the initial/observed value (in `( )`).

```{r}
#| label: print-lfmcmc-results
print(calibration_lfmcmc, burnin = lfmcmc_burnin)
```

Here is the posterior distribution of the LFMCMC samples with vertical lines representing the initial parameter values.

```{r}
#| label: plot-posterior-dist
plot_lfmcmc_post_dist(calibration_lfmcmc, init_lfmcmc_params, param_names, seasons)
```

## Epiworld Forecast

We can now run the forecast.
Our model prevalence is set according to the reported case counts of the most recent day of the UDHHS data.
We then take a sample of n = 200 from the LFMCMC accepted parameters (after the burn-in period) and run the SIR connected model with the new prevalence for each set of parameters.
Each simulation is for two weeks, giving us a 14-day forecast of COVID-19 in Utah.
The forecast mean is shown below along with the 50% and 95% confidence intervals.

```{r}
#| label: plot-forecast
plot_forecast(forecast_dist, covid_data)
```
